Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-01 LICS21/ranking.tex

\begin{example}[Geometric distribution]
% Consider the term
% \begin{align*}
% &(\tY \lambda f, n. \\
% &\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
% &) \, \underline{0}
% \end{align*}
%\[
%\big(\tY \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
%\big) \, \underline{0}
%\]
Let
\[
\Theta := \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)}. 
\]
The term  $(\tY \, \Theta) \, \underline 0$ generates a geometric distribution.
\changed[akr]{Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}.}
\lo{\Cref{fig:geometric distribution} is confusing: the arrow means either one-step reduction, or its reflexive transitive closure, or ``has the same syntactic shape''.}

@@

\begin{figure}
\begin{tikzpicture}[node distance=0.5em, anchor=west]
    \node (1) at (0,-0) {$\tY\, \Theta\, \underline 0$};
    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline i$};
    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})})\, \underline i$};
    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (6) at (0,-5) {$\tif{\underline r < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (7) at (2,-6) [text width=] {$\underline i$};
    \node (8) at (0,-7) {$\tY\, \Theta\, (\underline i\, \underline{+ 1})$};
    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};

    \draw [->] (1.south west) to [bend right=45] (2.north west);
    \draw [->] (2.south west) to [bend right=45] (3.north west);
    \draw [->] (3.south west) to [bend right=45] (4.north west);
    \draw [->] (4.south west) to [bend right=45] (5.north west);
    \draw [->] (5.south west) to [bend right=45] (6.north west);
    \draw [->] (6.south west) to [bend right=45] (8.north west);
    \draw [->] (8.south west) to [bend right=45] (9.north west);

    \draw [->] (9.west) to [bend left=45] (2.west);
    \draw [->] (6) to (7);
\end{tikzpicture}
\caption{Some reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution}}
\end{figure}
\begin{figure}
\centering
\begin{tikzpicture}[node distance=0.5em, anchor=west]
    \node (1) at (0,-0) {$\tY\, \Theta\, \underline i$ \quad [provided $i = 0$]};
    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline i$};
    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})})\, \underline i$};
    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (6) at (0,-5) {$\tif{\underline r' < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (7) at (2,-6) [text width=] {$\underline i$};
    \node (8) at (0,-7) {$\tY\, \Theta\, (\underline i\, \underline{+ 1})$};
    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
    \node (10) at (0,-9) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline{i + 1}$};

>>>>

\begin{example}[Geometric distribution]
% Consider the term
% \begin{align*}
% &(\tY \lambda f, n. \\
% &\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
% &) \, \underline{0}
% \end{align*}
%\[
%\big(\tY \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
%\big) \, \underline{0}
%\]
Let
\[
\Theta := \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)}. 
\]
The term  $(\tY \, \Theta) \, \underline 0$ generates a geometric distribution.
\changed[akr]{Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.}
\lo{\Cref{fig:geometric distribution} is confusing: the arrow means either one-step reduction, or its reflexive transitive closure, or ``has the same syntactic shape''.}

\akr{The arrows all mean the one-step reduction $\to$, I just didn't intend for the values of $i$ and $r$ to be consistent between the terms. I would much rather have the diagram be complete (with all the reductions and all the terms).}

@@

\begin{figure}
\begin{tikzpicture}[node distance=0.5em, anchor=west]
    \node (1) at (0,-0) {$\tY\, \Theta\, \underline 0$};
    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline i$};
    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})})\, \underline i$};
    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (6) at (0,-5) {$\tif{\underline {r - 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (7) at (2,-6.3) [text width=] {$\underline i$};
    \node (8) at (0,-7) {$\tY\, \Theta\, (\underline i\, \underline{+ 1})$};
    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};

    \draw [->] (1.south west) to [bend right=45, "$i := 0$"'] (2.north west);
    \draw [->] (2.south west) to [bend right=45] (3.north west);
    \draw [->] (3.south west) to [bend right=45] (4.north west);
    \draw [->] (4.south west) to [bend right=45] (5.north west);
    \draw [->] (5.south west) to [bend right=45] (6.north west);
    \draw [->] (6.south west) to [bend right=45, "$r \geq 0.5$"] (8.north west);
    \draw [->] (8.south west) to [bend right=45] (9.north west);

    \draw [->] (9.west) to [bend left=45, "$i := i + 1$" {above,sloped}] (2.west);
    \draw [->] (6) to ["$r < 0.5$"] (7);
\end{tikzpicture}
\caption{The reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution}}
\end{figure}
\begin{figure}
\centering
\begin{tikzpicture}[node distance=0.5em, anchor=west]
    \node (1) at (0,-0) {$\tY\, \Theta\, \underline i$ \quad [provided $i = 0$]};
    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline i$};
    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})})\, \underline i$};
    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (6) at (0,-5) {$\tif{\underline r' < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (7) at (2,-6) [text width=] {$\underline i$};
    \node (8) at (0,-7) {$\tY\, \Theta\, (\underline i\, \underline{+ 1})$};
    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
    \node (10) at (0,-9) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline{i + 1}$};
